---
title: "Welcome to STA 702"
subtitle: "Course Overview"
author: "Merlise Clyde"
institute: "Duke University"

format: 
  revealjs:
    theme: [simple, custom.scss]
    slide-number: true
    incremental: true
    preview-links: auto
    logo: ../img/icon.png
    footer: <https://sta702-F23.github.io/website/>
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


## What is this course about?

- Learn the foundations and theory of Bayesian inference in the context of several models.

- Use Bayesian models to answer inferential questions.

- Apply the models to several different problems.

-  Understand the advantages/disadvantages of Bayesian methods vs classical methods


. . .

 <i class="fa fa-quote-left fa-2x fa-pull-left fa-border" aria-hidden="true"></i>

<i class="fa fa-quote-right fa-2x fa-pull-right fa-border" aria-hidden="true"></i>
A Bayesian version will usually make things better...

-- Andrew Gelman.

## Instructional Team  

*Instructor:* [Dr Merlise Clyde](https://www2.stat.duke.edu/~clyde)

<i class="fa fa-envelope"></i> &nbsp; [clyde@duke.edu](mailto:clyde@duke.edu) <br>
<i class="fa fa-home"></i> &nbsp; 223 Old Chemistry <br>
<i class="fa fa-home"></i> &nbsp; [https://www2.stat.duke.edu/~clyde](https://www2.stat.duke.edu/~clyde) <br>


. . .

&nbsp;  

*Teaching Assistant:* [Rick Presman](https://scholars.duke.edu/person/rick.presman)

<i class="fa fa-envelope"></i> &nbsp; [rick.presman@duke.edu](mailto:rick.presman@duke.edu) 

&nbsp;   

<i class="fa fa-calendar"></i> &nbsp; See course website for Office Hours, Policies and more!


## Prerequisites  
 
 - random variables, common families of probability distribution functions and expectations
 - conditional distributions
 - transformations of random variables and change of variables 
 - principles of statistical inference (likelihoods) 
 - sampling distributions and hypothesis testing
 - concepts of convergence

. . .

Review Chapters 1 to 5 of the [Casella and Berger book](https://mybiostats.files.wordpress.com/2015/03/casella-berger.pdf)

## Computing

- Labs/HW will involve a lot of computing in R!

- Write your own MCMC samplers and run code long enough to show convergence

- You can learn R on the fly
  + see [Resources Tab on website](../resources.html)
  + materials from 2023 Bootcamp/Orientation
 
## Grading Policies
 
 
  + 5% class
  + 20% HW
  + 10% Lab
  + 20% Midterm I
  + 20% Midterm II
  + 25% Final

- No Late Submissions for HW/Lab; Drop the lowest score

- Confirm that you have access to Sakai, Gradescope, and GitHub.


 

