---
title: "STA 702: Lecture 0"
subtitle: "Course Overview"
author: "Merlise Clyde"
date: " "
format:
  html:
    self-contained: true
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

class: center, middle

# Welcome to STA 702!



---

## What is this course about?

<i class="fa fa-book fa-2x"></i> Learn the foundations and theory of Bayesian inference in the context of several models.

--




<i class="fa fa-folder-open fa-2x"></i> Apply the models to several different problem sets.


--

<i class="fa fa-tasks fa-2x"></i> Use Bayesian models to answer inferential questions.

--

<i class="fa fa-database fa-2x"></i> Understand the advantages/disadvantages of Bayesian methods vs classical methods


--

<i class="fa fa-clock fa-2x"></i> We will use the Hoff book, BDA and other reference material.

--

--- 
<i class="fa fa-quote-left fa-2x fa-pull-left fa-border" aria-hidden="true"></i>
<i class="fa fa-quote-right fa-2x fa-pull-right fa-border" aria-hidden="true"></i>
A Bayesian version will usually make things better...

-- Andrew Gelman.



---
## Instructor

[Dr Merlise Clyde](https://www2.stat.duke.edu/~clyde)

<i class="fa fa-envelope"></i> &nbsp; [clyde@duke.edu](mailto:clyde@duke.edu) <br>
<i class="fa fa-home"></i> &nbsp; [https://www2.stat.duke.edu/~clyde](https://www2.stat.duke.edu/~clyde) <br>
<i class="fa fa-university"></i> &nbsp; 223 Old Chemistry <br>
<i class="fa fa-home"></i> &nbsp; [https://www2.stat.duke.edu/courses/Fall22/sta702.001](https://www2.stat.duke.edu/courses/Fall22/sta702.001) <br>
<i class="fa fa-calendar"></i> &nbsp; See course website for OH <br>



---
## TAs

[Steven Winter](https://scholars.duke.edu/person/steven.winter)

<i class="fa fa-envelope"></i> &nbsp; [steven.winter@duke.edu](mailto:steven.winter@duke.edu) <br>
<i class="fa fa-calendar"></i> &nbsp; See course website for OH <br>
<i class="fa fa-university"></i> &nbsp; See course website for location <br>




---

## FAQs

All materials and information will be posted on the course webpage:
[https://www2.stat.duke.edu/courses/Fall22/sta702.001/](https://www2.stat.duke.edu/courses/Fall22/sta702.001/)

--

- How much theory will this class cover? A lot! 
--

- What if I can't remember all the topics in the prerequisites?

--
  1. Review Chapters 1 to 5 of the [Casella and Berger book](https://mybiostats.files.wordpress.com/2015/03/casella-berger.pdf)
     You can find the solution manual [here](http://www.ams.sunysb.edu/~zhu/ams570/Solutions-Casella-Berger.pdf)
 
--
 
  2. Focus on the following topics:
      + basic probability theory, random variables, transformations of random variables and change of variables, expectations of random variables, common families of probability distribution functions including multivariate distributions
      + concepts of convergence, principles of statistical inference, likelihood based inference, sampling distributions and hypothesis testing.



---

## FAQs


- Will we be doing "very heavy" computing? a good amount. You will be expected to be able to write your own MCMC samplers and run code long enough to show convergence

--

- What computing language will we use? R!

--

- What if I don't know R? This course assumes you already know R but you can still learn on the fly (you must be self-motivated). Here are some resources for you:
   + [https://www2.stat.duke.edu/courses/Fall22/sta702.001/resources/](https://www2.stat.duke.edu/courses/Fall22/sta702.001/resources/).
   + Resources for the StaSci BootCamp 2021

--

- Labs will introduce/review concepts

---
## FAQs

- Do I need to know a lot Bayesian statistics before taking this class? No

--

-  What is the difference between this course and STA360 or STA602 ?


---

class: center, middle

# Course structure and policies


---
## Course structure and policies

- See: [https://www2.stat.duke.edu/courses/Fall22/sta702.001/policies/](https://www2.stat.duke.edu/courses/Fall22/sta702.001/policies/)

--

- Make use of the teaching team's office hours, we're here to help!

--

- Do not hesitate to come to my office hours or you can also make an appointment to discuss a homework problem or any aspect of the course. 

--

- When the teaching team has announcements for you we will send an email to your Duke email address. Please make sure to check your email daily.

--

- Check the dates for exams in the Course Calendar and let me know ASAP if there are issues

---
## Topics

-  Basics of Bayesian Models
-  Loss Functions, Inference and Decision Making
-  Predictive Distributions
-  Predictive Distributions and Model Checking
-  Bayesian Hypothesis Testing
-  Multiple Testing
-  MCMC (Gibbs & Metropolis Hastings Algorithms)
-  Model Uncertainty
-  Bayesian Generalized Linear Models
-  Hiearchical Modeling and Random Effects
-  Hamiltonian Monte Carlo
-  Nonparametric Bayes Regression

--

Also refer to the  [Class Schedule](https://www2.stat.duke.edu/courses/Fall22/sta-702.001/syllabus/).

---
## Other details

- Grading 
  + 5% class
  + 20% HW
  + 10% Lab
  + 20% Midterm I
  + 20% Midterm II
  + 25% Final

--

- No Late Submissions for HW/Lab; Drop the lowest score

--

- Confirm that you have access to Sakai, Gradescope, and GitHub.

---
## Important Dates


<i class="fa fa-calendar"></i>  Tues, Aug 30  &nbsp; Classes begin <br>
<i class="fa fa-calendar"></i>  Fri, Sept 9   &nbsp; Drop/Add ends <br>
<i class="fa fa-calendar"></i>  Fri, Oct 7  &nbsp; Midterm I  (*tentative*) <br>
<i class="fa fa-calendar"></i>  Sat - Tues, Oct 8 - 11	&nbsp; No classes Fall Break |<br>
<i class="fa fa-calendar"></i> Tue, Nov 15    &nbsp; Midterm II (*tentative*) <br>
<i class="fa fa-calendar"></i>  Friday Dec 2    &nbsp;Graduate Classes End <br>
<i class="fa fa-calendar"></i>  Mon, Dec 19    &nbsp; Final Exam (Perkins 060 2:00-5:00pm) <br>

--

Also refer to the schedule on the website for updated breakdown of the courses. Remember to refresh the page frequently. See here: [Class Schedule](https://www2.stat.duke.edu/courses/Fall21/sta-601.001/syllabus/).

---

class: center, middle

# Bayes Rules!  Getting Started!


---
## Basics of Bayesian inference

- Generally (unless otherwise stated), in this course, we will use the following notation. Let

--

  + $\mathcal{Y}$ be the .hlight[sample space];

--

  + $y$ be the .hlight[observed data];

--

  + $\Theta$ be the .hlight[parameter space]; and

--

  + $\theta$ be the .hlight[parameter of interest].
  
--

- More to come later.


---
## Frequentist inference

- Given data $y$, estimate the population parameter $\theta$.

--

- How to estimate $\theta$ under the frequentist paradigm?
  + Maximum likelihood estimate (MLE)
  
  + Method of moments
  
  + and so on...
  
--

- Frequentist ML estimation finds the one value of $\theta$ that maximizes the likelihood.

--

- Typically uses large sample (asymptotic) theory to obtain confidence intervals and do hypothesis testing.


---
## What are Bayesian methods?

- .hlight[Bayesian methods] are data analysis tools derived from the principles of Bayesian inference and provide

--

  + parameter estimates with good statistical properties;
  
--

  + parsimonious descriptions of observed data;
  
--

  + predictions for missing data and forecasts of future data; and
  
--

  + a computational framework for model estimation, selection, decision making
and validation.

--
 
  + builds on likelihood inference


---
## Bayes' theorem - basic conditional probability

- Let's take a step back and quickly review the basic form of Bayes' theorem.

--

- Suppose there are some events $A$ and B having probabilities $\Pr(A)$ and $\Pr(B)$.

--

- Bayes' rule gives the relationship between the marginal probabilities of A and B and the conditional probabilities.

--

- In particular, the basic form of .hlight[Bayes' rule] or .hlight[Bayes' theorem] is
.block[
.small[
$$\Pr(A | B) = \frac{\Pr(A \ \textrm{and} \ B)}{\Pr(B)} = \frac{\Pr(B|A)\Pr(A)}{\Pr(B)}$$
]
]

--
$\Pr(A)$ = marginal probability of event $A$, $\Pr(B | A)$ = conditional probability of event $B$ given event $A$, and so on.


---
## Bayes' Rule more generally

  1. For each $\theta \in \Theta$, specify a .hlight[prior distribution] $p(\theta)$ or $\pi(\theta)$, describing our beliefs about $\theta$ being the true population parameter.

--

  2. For each $\theta \in \Theta$ and $y \in \mathcal{Y}$, specify a .hlight[sampling distribution] $p(y|\theta)$, describing our belief that the data we see $y$ is the outcome of a study with true parameter $\theta$. $p(y|\theta)$ proportional to the .hlight[likelihood] $L(\theta|y)$.

--

  3. After observing the data $y$, for each $\theta \in \Theta$, update the prior distribution to a .hlight[posterior distribution] $p(\theta | y)$ or $\pi(\theta | y)$, describing our "updated" belief about $\theta$ being the true population parameter.

--

- Now, how do we get from Step 1 to 3? .hlight[Bayes' rule]!
.block[
.small[
$$p(\theta | y) = \frac{p(\theta)p(y|\theta)}{\int_{\Theta}p(\tilde{\theta})p(y| \tilde{\theta}) \textrm{d}\tilde{\theta}} = \frac{p(\theta)p(y|\theta)}{p(y)}$$
]
]

  We will use this over and over throughout the course!

  
  
---
## Notes on prior distributions

Many types of priors may be of interest. These may

  + represent our own beliefs; 

--

  + represent beliefs of a variety of people with differing prior opinions; or

--

  + assign probability more or less evenly over a large region of the parameter space.
  
--

  + and so on...
  

---
## Notes on prior distributions

- .hlight[Subjective Bayes]: a prior should accurately quantify some individual's beliefs about $\theta$.

--

- .hlight[Objective Bayes]: the prior should be chosen to produce a procedure with "good" operating characteristics without including subjective prior knowledge.

--

- .hlight[Weakly informative]: prior centered in a plausible region but not overly-informative, as there is a tendency to be over
confident about one's beliefs.

--

-- .hlight[Practical Bayes]:  Combines Subjective Bayes for aspects of a problem that one understands, and Objective Bayes elsewhere
---
## Notes on prior distributions

- The prior quantifies 'your' initial uncertainty in $\theta$ before you observe new data (new information) - this may be necessarily subjective & summarize experience in a field or
prior research.

--

- Even if the prior is not "perfect", placing higher probability in a ballpark of the truth leads to better performance.

--

- Hence, it is very seldom the case that a weakly informative prior is not preferred over no prior.  (Model selection is one case where one needs to be careful!)

--

- One (very important) role of the prior is to stabilize estimates  (shrinkage) in the presence of limited data.

  
---
## Next Steps

Work on [Lab 0](https://www2.stat.duke.edu/courses/Fall22/sta702.001/syllabus/)

Finally, here are some readings to entertain you. Make sure to glance through them within the next week. See here: [Course Resources](https://www2.stat.duke.edu/courses/Fall22/sta702.001/resources) 
  1. Efron, B., 1986. Why isn't everyone a Bayesian?. The American Statistician, 40(1), pp. 1-5.
  2. Gelman, A., 2008. Objections to Bayesian statistics. Bayesian Analysis, 3(3), pp. 445-449.
  3. Diaconis, P., 1977. Finite forms of de Finetti's theorem on exchangeability. Synthese, 36(2), pp. 271-281.
  4. Gelman, A., Meng, X. L. and Stern, H., 1996. Posterior predictive assessment of model fitness via realized discrepancies. Statistica sinica, pp. 733-760.
  5. Dunson, D. B., 2018. Statistics in the big data era: Failures of the machine. Statistics & Probability Letters, 136, pp. 4-9.




