---
title: "The Normal Model & Prior/Posterior Predictive Distributions"
subtitle: "STA 702: Lecture 3"
author: "Merlise Clyde"
institute: "Duke University"
format: 
  revealjs:
    theme: [simple, custom.scss]
    slide-number: true
    incremental: true
    scrollable: false
    controls: true
    fragments: true
    preview-links: auto
    logo: ../../img/icon.png
    footer: <https://sta702-F23.github.io/website/>
    chalkboard: 
      boardmarker-width: 1
      chalk-width: 2
      chalk-effect: 0
    embed-resources: false
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"    
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: false
---


```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

## Outline


- Normal Model


- Predictive Distributions


 - Prior Predictive; useful for prior elicitation 
  

- Posterior Predictive; predicting/forecasting future events
  

- Comparing Estimators
  
 
## Normal Model Setup {.smaller}

- Suppose we have independent observations 
$$\mathbf{y} = (y_1,y_2,\ldots,y_n)^T$$ where each $Y_i  \mid \theta, \sigma^2 \stackrel{iid}{\sim} \textsf{N}(\theta, \sigma^2)$ 

  

- We will see that it is more convenient to work with $\tau = 1/\sigma^2$   (precision)


- reparameterizing the model for the data we have
$$Y_i \mid \theta, \tau  \sim \mathcal{N}(\theta, \tau^{-1})$$
  

- for simplicity  we will treat $\tau$ as known initially.  



- Need to specify a prior for $\theta$ on $\mathbb{R}$ 
 

## Prior for a Normal Mean {.smaller}

- Natural choice is a Normal/Gaussian distribution  (Conjugate prior)
$$\theta \sim \textsf{N}(\theta_0, 1/\tau_0)$$
  

- $\theta_0$ is the prior mean - best guess for $\theta$ using information other than $\mathbf{y}$

  
- $\tau_0$ is the prior precision and expresses our certainty about this guess

  

- one notion of non-informative is to let $\tau_0 \to 0$
 
  

- better justification is as Jeffreys' prior (uniform measure)  
$\pi(\theta) \propto 1$

- parameterization invariant and invariant to location/scale changes in the data  (group invariance)

. . .

::: {.callout-important title="Exercise for the Energetic Student"}
You should be able to derive Jeffreys prior!
:::
 
## Posterior Distribution (1 observaton) {.smaller}

- Posterior  $$p(\theta \mid y) \propto \exp\left\{- \frac 1 2 [\tau (y - \theta) ^2 + \tau_0(\theta - \theta_0) ^2 \right\} \, d\theta$$

- Quadratic in exponential term:
$\tau_0(\theta - \theta_0)^2 = \tau_0 \theta^2 - 2 \tau_0 \theta_0 \theta + \tau_0 \theta_0^2$

    - Expand quadratics, regroup and read off precision from  quadtric term in $\theta$ and mean from linear term in $\theta$



- posterior precision is the sum of prior precision and data precision $\tau_0 + \tau$


- posterior mean $\hat{\theta} = \frac{\tau_0} {\tau_0 + \tau} \theta_0 + \frac{\tau}{\tau_0 + \tau} y$; precision weighted average of prior mean and MLE

  



 - conjugate family updating
$\theta \mid y \sim \textsf{N} \left(\hat{\theta}, \frac{1}{\tau_0 + \tau} \right)$

 
## Marginal  Distribution {.smaller}

- Recall that the [**marginal distribution**]{style="color:red"} is
$$p({y}) = p(y_1,\ldots,y_n) = \int_\Theta p(y_1,\ldots,y_n \mid \theta) \pi(\theta)\, d\theta$$

  

- this is also called the [**prior predictive**]{style="color:red"} distribution and is independent of any unknown parameters


- We may care about making predictions before we even see any data.

  

- This is often useful as a way to see if the sampling distribution or prior we have chosen is appropriate, after integrating out all unknown parameters.


  



  
 
## Prior Predictive  for a Single Case {.smaller}

$$\begin{split} p(y) & \propto \int_\mathbb{R} p(y \mid \theta) \pi(\theta) \, d\theta \\
& \propto \int_\mathbb{R}\exp\left\{- \frac 1 2 \tau (y - \theta) ^2 \right\} \exp\left\{-  \frac 1 2 \tau_0(\theta - \theta_0) ^2 \right\} \, d\theta
\end{split}$$

 . . . 


  
:::: {.columns}
::: {.column width="50%"}
Integration

1)  [**Expand**]{style="color:red"} quadratics  in exp terms

  

2)  [**Group**]{style="color:red"} terms with $\theta^2$ and $\theta$

  

3) Read off  [**posterior precision**]{style="color:red"} and  
[**posterior mean**]{style="color:red"}

  

4)  [**Complete the square**]{style="color:red"} 

  

5)  [**Integrate**]{style="color:red"} out $\theta$ to obtain marginal!

:::



::: {.column width="50%"}
1) Linear combinations of Normals are Normal!
$$Y \stackrel{D}{=} \theta + \epsilon, \quad \epsilon \sim N(0, 1/\tau) \quad \theta \sim N(\theta_0, 1/\tau_0)$$

2) Find Mean of sum

3) Find Variance of sum

4) Marginal $Y \sim  N(\theta_0, 1/\tau_0 + 1/\tau)$
:::

::::

## Prior Predictive {.smaller}

- marginal distribution for $Y$ (prior predictive)
$$Y \sim \textsf{N}\left(\theta_0, \frac{1}{\tau_0} + \frac{1}{\tau}\right) \text{ or } \textsf{N}(\theta_0, \sigma^2 + \sigma^2_0)$$

  

- two sources of variability:  data variability  and prior variability




- useful to think about observable quantities when choosing the prior

  

- sample directly  from the prior predictive and assess whether the samples are consistent with our prior knowledge

  

- if not, go back and  modify the prior & repeat

  

- sequential substitution sampling  (repeat $T$ times)

  1) draw $\theta^{(t)} \sim \pi(\theta)$

  2) draw $y^{(t)} \mid \theta^{(t)} \sim p(y \mid \theta^{(t)})$


  

- takes into account uncertain about $\theta$ and variability in $Y$!


 
## Posterior Updating {.smaller}



- Sequential updating using the previous result as our prior!

  

- New prior after seeing 1 observation is $$\textsf{N}(\theta_1, 1/\tau_1)$$

  

- prior mean weighted average
  $$\theta_1 \equiv \frac{\tau_0 \theta_0 + \tau y_1}{\tau_0 + \tau_1}$$
  
  

- prior precision after 1 observation
  $$\tau_1 \equiv \tau_0 + \tau$$

  


- prior variance is now $\sigma^2_1 = 1/\tau_1$

 
## Posterior Predictive for $y_2$ given $y_1$  {.smaller}

- Conditional $p(y_2 \mid y_1) = p(y_2, y_1)/p(y_1)$   (Hard way!)

  

- Use latent variable representation
$$p(y_2 \mid y_1) = \int_\Theta \frac{p(y_2, \mid \theta) p( y_1 \mid \theta ) \pi(\theta) \, d\theta}{p(y_1)}$$
  

- simplify to previous problem and use results
$$p(y_2 \mid y_1) =  \int_\Theta p(y_2 \mid \theta) \pi(\theta \mid y_1) \, d\theta$$
  

- (Posterior) Predictive
$$Y_2 \mid y_1 \sim \textsf{N}(\theta_1, \sigma^2 + \sigma^2_1)$$
 
## Iterated Expectations  {.smaller}

Based on expressions we have an exponential of a quadratic in $y_2$ so know that distribution is Gaussian

  

- Find the mean and variance using iterated expectations:

  

- mean 
$$\textsf{E}[Y_2 \mid y_1] = \textsf{E}_{\theta \mid y_1}[\textsf{E}_{Y_2 \mid y_1, \theta} [Y_2 \mid y_1, \theta] \mid y_1]$$

- Conditional Variance $\textsf{Var}[Y_2  \mid y_1]$

- Iterated expectations (prove!)
$$\textsf{Var}[Y_2  \mid y_1] = \textsf{E}_{\theta \mid y_1}[\textsf{Var}_{Y_2 \mid y_1, \theta} [Y_2 \mid y_1, \theta] \mid y_1] + \textsf{Var}_{\theta \mid y_1}[\textsf{E}_{Y_2 \mid y_1, \theta} [Y_2 \mid y_1, \theta] \mid y_1]$$


 
## Updated Posterior for $\theta$ {.smaller}

$$p(\theta \mid y_1, y_2) \propto p(y_2 \mid \theta) p(y_1 \mid \theta) \pi(\theta)$$

  

$$p(\theta \mid y_1, y_2) \propto  p(y_2 \mid \theta) p(\theta \mid y_1)$$

. . .


Apply previous updating rules

  

- new posterior mean 
$$\theta_2 = \frac{\tau_1 \theta_1  + \tau y_2}{\tau_1 + \tau} = \frac{\tau_0 \theta_0 + 2 \tau \bar{y}}
{\tau_0 + 2 \tau}$$
  

- new precision
$$ \tau_2 = \tau_1 + \tau = \tau_0 + 2 \tau$$

 
## After $n$ observations  {.smaller}

- Posterior for $\theta$ 
$$\theta \mid y_1, \ldots, y_n \sim \textsf{N}\left( \frac{\tau_0 \theta_0 + n \tau \bar{y}}
{\tau_0 + n \tau}, \frac{1}{ \tau_0 + n \tau} \right)$$

  

- Posterior Predictive Distribution for $Y_{n+1}$
$$Y_{n+1} \mid y_1, \ldots, y_n \sim \textsf{N}\left( \frac{\tau_0 \theta_0 + n \tau \bar{y}}
{\tau_0 + n \tau}, \frac{1}{\tau} + \frac{1}{ \tau_0 + n \tau} \right)$$

  

- Shrinkage of the MLE to the prior mean

  

- More accurate estimation of $\theta$ as $n \to \infty$ (reducible error)

  

- Cannot reduce the error for prediction $Y_{n+1}$ due to $\sigma^2$

  

- predictive distribution for a next observation given _everything_ we know - prior and likelihood

 
## Results with Jeffreys' Prior {.smaller}
 - What if $\tau_0 \to 0$? (or $\sigma^2_0 \to \infty$)
 
  

- Prior predictive $\textsf{N}(\theta_0, \sigma^2_0 + \sigma^2 )$  (not  proper in the limit)

  

- Posterior for $\theta$ (formal posterior)
$$\theta \mid y_1, \ldots, y_n \sim \textsf{N}\left( \frac{\tau_0 \theta_0 + n \tau \bar{y}}
{\tau_0 + n \tau}, \frac{1}{ \tau_0 + n \tau} \right)$$

 . . . 

$$\to  \qquad \theta \mid y_1, \ldots, y_n \sim \textsf{N}\left( \bar{y}, 
 \frac{1}{n \tau} \right)$$
  

- Recovers the MLE as the posterior mode!

  

- Posterior variance of $\theta = \sigma^2/n$ (same as variance of the MLE)
 

 
## Posterior Predictive Distribution {.smaller}

- Posterior predictive distribution for $Y_{n+1}$
$$Y_{n+1} \mid y_1, \ldots, y_n \sim \textsf{N}\left( \frac{\tau_0 \theta_0 + n \tau \bar{y}}
{\tau_0 + n \tau}, \frac{1}{\tau} + \frac{1}{ \tau_0 + n \tau} \right)$$

  

- Under Jeffreys' prior
$$Y_{n+1} \mid y_1, \ldots, y_n \sim \textsf{N}\left( \bar{y}, \sigma^2 (1 + \frac{1}{n} )\right)$$
  

- Captures extra uncertainty due to not knowing $\theta$ (compared to plug-in approach where we plug in MLE in sampling model!

 
## Comparing Estimators {.smaller}

- Expected loss (from frequentist perspective) of using Bayes Estimator

  

- Posterior mean is optimal under squared error loss (min Bayes Risk)  [also absolute error loss]

  

- Compute Mean Square Error (or Expected Average Loss)
$$\textsf{E}_{\bar{y} \mid \theta}\left[\left(\hat{\theta} - \theta \right)^2 \mid \theta \right]$$

 . . .

 $$ = \textsf{Bias}(\hat{\theta})^2 + \textsf{Var}(\hat{\theta})$$
  

- For the MLE $\bar{Y}$ this is just the variance of $\bar{Y}$ or $\sigma^2/n$

 
## MSE for Bayes {.smaller}

- Frequentist Risk
$$\textsf{E}_{\bar{y} \mid \theta}\left[\left(\hat{\theta} - \theta \right)^2 \mid \theta \right] = \textsf{MSE} =  \textsf{Bias}(\hat{\theta})^2 + \textsf{Var}(\hat{\theta})$$

- Bias of Bayes Estimate
$$\textsf{E}_{\bar{Y} \mid \theta}\left[ \frac{\tau_0 \theta_0 + \tau n \bar{Y}}
{\tau_0  + \tau n}\right] = 
\frac{\tau_0(\theta_0 - \theta)}{\tau_0 + \tau n}$$

  

- Variance
$$\textsf{Var}\left(\frac{\tau_0 \theta_0 + \tau n \bar{Y}}{\tau_0 + \tau n} - \theta  \mid \theta \right)  = \frac{\tau n}{(\tau_0 + \tau n)^2}$$
  

- (Frequentist) expected Loss  when truth is $\theta$
$$\textsf{MSE} = \frac{\tau_0^2(\theta - \theta_0)^2 + \tau n}{(\tau_0 + \tau n)^2}$$
  



 
## Plot {.smaller}

Behavior ?
```{r MSE, fig.width=5, fig.height=5,  out.width="75%", echo=FALSE}
sigma = 1
n = 25

MSE = function(theta, tau=1, n, theta_0=0, tau_0 = .05) {
(tau_0^2*(theta - theta_0)^2 + tau*n)/(tau_0 + tau*n)^2
}

theta = seq(-10,10, length=10000)
plot(theta,  MSE(theta, tau=1/sigma^2, n = n), 
     type="l", col=2, lwd=2, 
     xlab=expression(theta),
     ylab="MSE")
abline(h=sigma^2/n, lty=2, col=1, lwd=2)
legend(x="center", legend=c("MLE", "Bayes"), col=c(1,2), lwd=2, lty=c(1,2))
```


 
## Updating with $n$ Observations {.smaller}

-  Can update sequentially as before -or-
 
 -  We can use the $\cal{L}(\theta)$ based on $n$ observations  and repeat completing the square with the original prior  $\theta \sim \textsf{N}(\theta_0, 1/\tau_0)$

- same answer!

- The likelihood for $\theta$ is proportional to the sampling model
$$p(y \mid \theta,\tau)  = 
\prod_{i=1}^n \frac{1}{\sqrt{2\pi}} \tau^{\frac{1}{2}}  \exp{\left\{-\frac{1}{2} \tau (y_i-\theta)^2\right\}}$$


 . . . 

::: {.callout-important title="Exercise"}
Rewrite in terms of sufficient statistics!
:::

## Exercises for Practice



:::: {.callout-important title="Exercise 1"}

Use $\cal{L}(\theta)$  based on $n$ observations and $\pi(\theta)$ to find  $\pi(\theta \mid y_1, \ldots, y_n)$ based on the sufficient statistics

:::
  
  . . .
  

::: {.callout-important title="Exercise 2"}

Use $\pi(\theta \mid y_1, \ldots, y_n)$ to find the posterior predictive distribution for $Y_{n+1}$

:::



 
## Simplification {.smaller}


$$\begin{split}
\cal{L}(\theta) & \propto \tau^{\frac{n}{2}} \ \exp\left\{-\frac{1}{2} \tau \sum_{i=1}^n (y_i-\theta)^2\right\}\\
& \propto \tau^{\frac{n}{2}} \ \exp\left\{-\frac{1}{2} \tau \sum_{i=1}^n \left[ (y_i-\bar{y}) - (\theta - \bar{y}) \right]^2 \right\}\\
\\
& \propto \tau^{\frac{n}{2}} \ \exp\left\{-\frac{1}{2} \tau \left[ \sum_{i=1}^n (y_i-\bar{y})^2 + \sum_{i=1}^n(\theta - \bar{y})^2 \right] \right\}\\
& \propto \tau^{\frac{n}{2}} \ \exp\left\{-\frac{1}{2} \tau \left[ \sum_{i=1}^n (y_i-\bar{y})^2 + n(\theta - \bar{y})^2 \right] \right\}\\
& \propto \tau^{\frac{n}{2}} \ \exp\left\{-\frac{1}{2} \tau s^2(n-1) \right\} \ \exp\left\{-\frac{1}{2} \tau n(\theta - \bar{y})^2 \right\}
\end{split}$$

 
