---
title: "Lecture 8: Metropolis-Hastings, Gibbs and Blocking"
subtitle: "STA702"
author: "Merlise Clyde"
institute: "Duke University"
format: 
  revealjs:
    theme: [simple, custom.scss]
    slide-number: true
    incremental: true
    scrollable: false
    controls: true
    fragments: true
    preview-links: auto
    logo: ../../img/icon.png
    footer: <https://sta702-F23.github.io/website/>
    chalkboard: 
      boardmarker-width: 1
      chalk-width: 2
      chalk-effect: 0
    embed-resources: false
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"    
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
require(lattice)
require(MCMCpack)
require(hdrcde)
require(coda)
require(mvtnorm)
#require(rethinking)
set.seed(42)
```


##  Metropolis-Hastings  (MH) {.smaller}

- Metropolis requires that the proposal distribution be symmetric

  

- Hastings (1970)  generalizes Metropolis algorithms to allow asymmetric proposals - aka Metropolis-Hastings or MH  $q(\theta^* \mid \theta^{(s)})$ does not need to be the same as $q(\theta^{(s)} \mid \theta^*)$

  
-  propose $\theta^*  \mid \theta^{(s)} \sim q(\theta^* \mid \theta^{(s)})$

  
- Acceptance probability 
$$\min \left\{ 1, \frac{\pi(\theta^*) \cal{L}(\theta^*)/q(\theta^* \mid \theta^{(s)})}
{\pi(\theta^{(s)}) \cal{L}(\theta^{(s)})/q( \theta^{(s)} \mid \theta^*)} \right\}$$

  

- adjustment for asymmetry in acceptance ratio is key to ensuring convergence to stationary distribution!

 
## Special cases {.smaller}

- Metropolis

  

- Independence chain

  

- Gibbs samplers

  

- Metropolis-within-Gibbs

  

- combinations of the above!

 
## Independence Chain {.smaller}

- suppose we have a good approximation $\tilde{\pi}(\theta \mid y)$   to  $\pi(\theta \mid y)$

  

-  Draw $\theta^* \sim \tilde{\pi}(\theta \mid y)$ _without_ conditioning on $\theta^{(s)}$

  

- acceptance probability 
$$\min \left\{ 1, \frac{\pi(\theta^*) \cal{L}(\theta^*)/\tilde{\pi}(\theta^* \mid \theta^{(s)})}
{\pi(\theta^{(s)}) \cal{L}(\theta^{(s)})/\tilde{\pi}( \theta^{(s)} \mid \theta^*)} \right\}$$

  

- what happens if the approximation is really accurate?

  

- probability of acceptance is $\approx 1$

  

- Important caveat for convergence:  tails of the posterior should be at least as heavy as the tails of the posterior  (Tweedie 1994)

  

-  Replace Gaussian by a Student-t with low degrees of freedom

  

- transformations of $\theta$



## Blocked Metropolis-Hastings {.smaller}

So far all algorithms update all of the parameters simultaneously

  

- convenient to break problems in to $K$ blocks and update them separately

  

- $\theta = (\theta_{[1]}, \ldots, \theta_{[K]}) = (\theta_1, \ldots, \theta_p)$



  

- At iteration $s$, for $k = 1, \ldots, K$ Cycle thru blocks: (fixed order or random order)

    - propose $\theta^*_{[k]} \sim q_k(\theta_{[k]} \mid \theta_{[<k]}^{(s)}, \theta_{[>k]}^{(s-1)})$
   
    - set $\theta_{[k]}^{(s)} = \theta^*_{[k]}$ with probability
   $$\min \left\{ 1, \frac{
   \pi(\theta_{[<k]}^{(s)},\theta_{[k]}^*, 
        \theta_{[>k]}^{(s-1)})
   \cal{L}(\theta_{[<k]}^{(s)},\theta_{[k]}^*,
           \theta_{[>k]}^{(s-1)})/
   q_k(\theta_{[k]}^* \mid \theta_{[<k]}^{(s)},    
       \theta_{[>k]}^{(s-1)})}
  {\pi(\theta_{[<k]}^{(s)},\theta_{[k]}^{(s-1)}, 
        \theta_{[>k]}^{(s-1)})
   \cal{L}(\theta_{[<k]}^{(s)},\theta_{[k]}^{(s-1)},
           \theta_{[>k]}^{(s-1)})/
   q_k(\theta_{[k]}^{(s-1)} \mid \theta_{[<k]}^{(s)},    
       \theta_{[>k]}^{(s-1)})} \right\}$$
 
## Gibbs Sampler {.smaller}

- The Gibbs Sampler is special case of Blocked MH



- proposal distribution $q_k$ for the $k$th block is the **full conditional** distribution for $\theta_{[k]}$
$$\begin{split}
\pi(\theta_{[k]} \mid \theta_{[-k]}, y) & = \frac{\pi(\theta_{[k]} , \theta_{[-k]} \mid y)}{ \pi(\theta_{[-k]} \mid y))} \propto \pi(\theta_{[k]} , \theta_{[-k]} \mid y)\\
\   & \propto \cal{L}(\theta_{[k]} , \theta_{[-k]})\pi(\theta_{[k]} , \theta_{[-k]})
\end{split}$$
  

- Acceptance probability 
$$\min \left\{ 1, \frac{
   \pi(\theta_{[<k]}^{(s)},\theta_{[k]}^*, 
        \theta_{[>k]}^{(s-1)})
   \cal{L}(\theta_{[<k]}^{(s)},\theta_{[k]}^*,
           \theta_{[>k]}^{(s-1)})/
   q_k(\theta_{[k]}^* \mid \theta_{[<k]}^{(s)},    
       \theta_{[>k]}^{(s-1)})}
  {\pi(\theta_{[<k]}^{(s)},\theta_{[k]}^{(s-1)}, 
        \theta_{[>k]}^{(s-1)})
   \cal{L}(\theta_{[<k]}^{(s)},\theta_{[k]}^{(s-1)},
           \theta_{[>k]}^{(s-1)})/
   q_k(\theta_{[k]}^{(s-1)} \mid \theta_{[<k]}^{(s)},    
       \theta_{[>k]}^{(s-1)})} \right\}$$
       
  

- Simplifies so that acceptance probability is always 1!

  

- even though joint distribution is messy, full conditionals may be (conditionally) conjugate and easy to sample from!


 
##  Univariate Normal Example {.smaller}

Model
$$\begin{align*}
Y_i \mid \mu, \sigma^2 & \overset{iid}{\sim} \textsf{N}(\mu, 1/\phi) \\
\mu & \sim \textsf{N}(\mu_0, 1/\tau_0) \\
\phi & \sim  \textsf{Gamma}(a/2, b/2)
\end{align*}$$

  
-  Joint prior is a product of independent  Normal-Gamma

  
-  Is  $\pi(\mu, \phi \mid y_1, \ldots, y_n)$ also a Normal-Gamma family?



 
## Full Conditional for the  Mean {.smaller}

The full conditional distributions  $\mu \mid \phi, y_1, \ldots, y_n$ 
$$\begin{align*}
\mu & \mid \phi, y_1, \ldots, y_n \sim \textsf{N}(\hat{\mu}, 1/\tau_n) \\
\hat{\mu} & = \frac{\tau_0 \mu_0  + n \phi \bar{y}}{\tau_0 + n \phi} \\
\tau_n & = \tau_0 + n \phi
\end{align*}$$
  
  
  
 
## Full Conditional for the Precision {.smaller}

- Full conditional for $\phi$
$$\begin{align*}
\phi  \mid \mu, y_1, \ldots, y_n & \sim \textsf{Gamma}( a_n/2, b_n/2) \\
a_n & = a + n \\
b_n & = b + \sum_i (y_i - \mu)^2
\end{align*}$$

 . . . 

$$\textsf{E}[\phi \mid \mu, y_1, \ldots, y_n] = \frac{(a + n)/2}{(b + \sum_i (y_i - \mu)^2 )/2}$$

  
- What happens with a non-informative prior  i.e
$a = b = \epsilon$ as $\epsilon \to 0$?

 . . .
  
::: {.callout-warning}
Proper full conditionals  with improper priors do not ensure proper joint posterior!
:::
 
## Normal Linear Regression Example {.smaller}

- Model
$$\begin{align*}
Y_i \mid \beta, \phi & \overset{iid}{\sim} \textsf{N}(x_i^T\beta, 1/\phi) \\
Y \mid \beta, \phi & \sim \textsf{N}(X \beta, \phi^{-1} I_n) \\
\beta & \sim \textsf{N}(b_0, \Phi_0^{-1}) \\
\phi & \sim \textsf{N}(v_0/2, s_0/2)
\end{align*}$$

- $x_i$ is a $p \times 1$ vector of predictors and $X$ is $n \times p$ matrix

  

- $\beta$ is a $p \times 1$ vector of coefficients

  

- $\Phi_0$ is a $p \times p$ prior precision matrix 

  

- Multivariate Normal density for $\beta$
$$\pi(\beta \mid b_0, \Phi_0) = \frac{|\Phi_0|^{1/2}}{(2 \pi)^{p/2}}\exp\left\{- \frac{1}{2}(\beta - b_0)^T \Phi_0 (\beta - b_0)  \right\}$$
 
## Full Conditional for $\beta$ {.smaller}

$$\begin{align*}
\beta & \mid \phi, y_1, \ldots, y_n \sim \textsf{N}(b_n, \Phi_n^{-1}) \\
b_n & =  (\Phi_0 + \phi X^TX)^{-1}(\Phi_0 b_0  +  \phi X^TX \hat{\beta})\\
\Phi_n & = \Phi_0 + \phi X^TX
\end{align*}$$

 
## Derivation continued {.smaller}


 
## Full Conditional for $\phi$ {.smaller}

$$ \phi \mid \beta, y_1, \ldots, y_n \sim \textsf{Gamma}((v_0 + n)/2, (s_0 + \sum_i(y_i - x^T_i \beta)))$$



 
##  Choice of Prior Precision {.smaller}

- Non-Informative $\Phi_0 \to 0$ 

  

- Formal Posterior given $\phi$
  $$\beta \mid \phi, y_1, \ldots, y_n \sim \textsf{N}(\hat{\beta}, \phi^{-1} (X^TX)^{-1})$$
  
  

- needs $X^TX$ to be full rank for distribution to be unique

 
## Invariance and Choice of Mean/Precision {.smaller}

-  the model in vector form
$$ Y \sim \textsf{N}_n (X\beta, \phi^{-1} I_n)$$

  

- What if we transform the $X$ matrix by $\tilde{X} = X H$ where $H$ is $p \times p$ and invertible

  

- obtain the posterior for $\tilde{\beta}$ using $Y$ and $\tilde{X}$  
$$ Y \sim \textsf{N}_n (\tilde{X}\tilde{\beta}, \phi^{-1} I_n)$$

- since $\tilde{X} \tilde{\beta} = X H  \tilde{\beta} = X \beta$  invariance suggests that the posterior for $\beta$ and $H \tilde{\beta}$ should be the same 

  
- or the posterior of $H^{-1} \beta$
and $\tilde{\beta}$ should be the same

  

- with some linear algebra we can show that this is true if $b_0 = 0$ and $\Phi_0$ is $k X^TX$ for some $k$  (show!)

 
## Zellner's g-prior {.smaller}

- Popular choice is to take $k = \phi/g$ which is a special case of Zellner's g-prior
$$\beta \mid \phi, g \sim \textsf{N}\left(0, \frac{g}{\phi} (X^TX)^{-1}\right)$$

  

- Full conditional 
$$\beta \mid \phi, g \sim \textsf{N}\left(\frac{g}{1 + g} \hat{\beta}, \frac{1}{\phi} \frac{g}{1 + g} (X^TX)^{-1}\right)$$
  

- one parameter $g$ controls shrinkage

  

- if $\phi \sim \textsf{Gamma}(v_0/2, s_0/2)$ then posterior is
$$\phi \mid y_1, \ldots, y_n \sim \textsf{Gamma}(v_n/2, s_n/2)$$
  

- Conjugate so we could skip Gibbs sampling and sample directly from gamma and then conditional normal!

 
## Ridge Regression  {.smaller}

- If $X^TX$ is nearly singular, certain  elements of $\beta$ or (linear combinations of $\beta$) may have huge variances under the $g$-prior (or flat prior) as the MLEs are highly unstable!

  

- **Ridge regression** protects against the explosion of variances and ill-conditioning with the conjugate priors:
$$\beta \mid \phi \sim \textsf{N}(0, \frac{1}{\phi \lambda} I_p)$$
  

- Posterior for $\beta$  (conjugate case)
$$\beta \mid \phi, \lambda, y_1, \ldots, y_n \sim 
\textsf{N}\left((\lambda I_p + X^TX)^{-1} X^T Y,  \frac{1}{\phi}(\lambda I_p + X^TX)^{-1}
\right)$$




 
##  Bayes Regression {.smaller}

- Posterior mean (or mode) given $\lambda$ is biased, but can show that there **always** is a value of $\lambda$  where the frequentist's expected squared error loss is smaller for the Ridge estimator than MLE!

  

- related to penalized maximum likelihood estimation 

  

-  Choice of $\lambda$

  

-  Bayes Regression and choice of $\Phi_0$ in general is a very important problem and provides the foundation  for many variations on shrinkage estimators, variable selection, hierarchical models, nonparameteric regression and more!

  

- Be sure that you can derive the full conditional posteriors for $\beta$ and $\phi$ as well as the joint posterior in the conjugate case!


 
## Comments {.smaller}



- Why don't we treat each individual $\beta_j$ as a separate block?

  

-  Gibbs always accepts, but can mix slowly if parameters in different blocks are highly correlated!

  

- Use block sizes in Gibbs that are as big as possible to improve  mixing (proven faster convergence)

  

- Collapse the sampler by integrating out as many parameters as possible  (as long as resulting sampler has good mixing)

  

- can use Gibbs steps and (adaptive) Metropolis Hastings steps together

  


-  Introduce latent variables (data augmentation) to allow Gibbs steps  (Next class)





